Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2

### 2023-10-17 go3-toMPN.tex
sent for comment

\title{Beyond compositionally in high throughput sequencing; estimating
the importance of scale in data analysis with ALDEx2}
\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin Silverman$^{2,3}$}
\date{\today}

@@

Many groups have conducted benchmarking studies on different tools and
normalizations used for the analysis of datasets such as transcriptomes
(Bullard et al. 2010; Soneson and Delorenzi 2013; Schurch et al. 2016;
Quinn, Crowley, and Richardson 2018) and microbiomes(Thorsen et al.
2016; Weiss et al. 2017; Hawinkel et al. 2018). Generically, it is
observed that the actual agreement between methods can be modest, and
there is usually a recommendation that some methods are better suited
for individual datasets or for specific types of analysis;
e.g.~`Normalization and microbial differential abundance strategies
depend upon data characteristics' (Weiss et al. 2017). However, another
interpretation of these studies is that the specific assumptions of the
tools and the way they are used by different groups can lead to results
that include both false negative and false positive results, skewed
towards the latter. There are now published examples of multiple large
replication efforts in psychology (Open Science Collaboration 2015),
cancer biology (Rodgers and Collings 2021), ecology and evolution (Gould
and et.al, n.d.). In a nutshell, experiences from these studies indicate
that published data typically have higher effect sizes and appear more
significant than replication efforts with the same data, or with newly
generated replication data (Rodgers and Collings 2021). Thus, most
published data should be viewed with suspicion (Ioannidis 2005) unless
efforts have been made to ensure that the results are robust

@@

\leavevmode\vadjust pre{\hypertarget{ref-Ioannidis:2005aa}{}}%
Ioannidis, John P A. 2005. {``Why Most Published Research Findings Are
False.''} \emph{PLoS Med} 2 (8): e124.
\url{https://doi.org/10.1371/journal.pmed.0020124}.

>>>>

\title{Beyond compositionally in high throughput sequencing; estimating
the importance of scale in data analysis with ALDEx2}
\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin Silverman$^{2,3}$}

\date{\today}

@@

Many groups have conducted benchmarking studies on different tools and
normalizations used for the analysis of datasets such as transcriptomes
(Bullard et al. 2010; Soneson and Delorenzi 2013; Schurch et al. 2016;
Quinn, Crowley, and Richardson 2018) and microbiomes(Thorsen et al.
2016; Weiss et al. 2017; Hawinkel et al. 2018). Generically, it is
observed that the actual agreement between methods can be modest, and
there is usually a recommendation that some methods are better suited
for individual datasets or for specific types of analysis;
e.g.~`Normalization and microbial differential abundance strategies
depend upon data characteristics' (Weiss et al. 2017). However, another
interpretation of these studies is that the specific assumptions of the
tools and the way they are used by different groups can lead to results
that include both false negative and false positive results, skewed
towards the latter. There are now published examples of multiple large
replication efforts in psychology (Open Science Collaboration 2015),
cancer biology (Rodgers and Collings 2021), ecology and evolution (Gould
and et.al, n.d.). In a nutshell, experiences from these studies indicate
that published data typically have higher effect sizes and appear more
significant than replication efforts with the same data, or with newly
generated replication data (Rodgers and Collings 2021). This is if the
data can be replicated de-novo at all {[}Errington:2021aa{]}.

@@

### 2023-10-17 go3.Rmd
sent for comment

---
title: "Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2
"
shorttitle: "Scale ALDEx2"
author:
- name: Greg Gloor
  affiliation: Dep't of Biochemistry, University of Western Ontario
  email: ggloor@uwo.ca
- name: Michelle Pistner Nixon
  affiliation: College of Information Sciences and Technology, Pennsylvania  State University
- name: Justin Silverman
  affiliation: College of Information Sciences and Technology, Department of Statistics, Department of Medicine, Pennsylvania  State University
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
output:
  BiocStyle::pdf_document: default
package: ALDEx2
abstract: |
    Introduction to scale simulation and FDR correction with ALDEx2.
---

@@

High throughput sequencing (HTS) is a universal tool to explore many biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing). HTS proceeds by taking a fixed-number sample from the environment, making a library, multiplexing (merging) multiple libraries and applying a fixed number of molecules to the flow cell. In essence it is a poll of the environment that is mixed with other polls and then a poll of the mixture is taken. It should be clear that the total number of molecules sequenced is driven by the capacity of the instrument and not by the number of molecules in the sampled environment[@Lovell:2015;@Vandeputte:2017aa;@Props:2017aa].

@@

More formally, if the true information in the  environment being sampled can be summarized by counts of features (genes, taxa, etc) in a matrix \( \mathbf{W} \), then the elements of the matrix can be decomposed into its composition (relative information, \( \mathbf{W}^{\parallel} \) ) and its scale (total sum, \( \mathbf{W}^{\perp} \) ); formally, \(\mathbf{W} = \mathbf{W}^{\parallel} \times \mathbf{W}^{\perp}\). If the matrix has \(N\) samples and \(D\) parts, then we can uniquely identify \(\mathbf{W}_{dn}\) as the \(d^{th}\) feature in sample \(n\).  Any data that we collect by sequencing is thought to contain only proportional [@lovell:2011] information and we can represent the corresponding sequenced matrix as \(\mathbf{Y}\). Compositional data analysis makes the strong assumption  that \(\mathbf{Y}_{n} \equiv \mathbf{Y}^{\parallel}_{n} = f(\mathbf{W}^{\parallel}_{n})\); in other words that what is measured by sequencing is relative information only, and that this measurement is a function of the relative information in the underlying environment. Note that under this  assumption no corresponding estimate or information is available for the scale of the sample \(\mathbf{Y}^{\perp}_{n}\).

@@

Recently, Nixon et al. [-@nixon2023scale] showed that that *all* normalizations make some  identifying assumption about the scale of \(\mathbf{Y}\) but that the actual assumption being made was not always easy to interpret. The CLR which can be written as \( \mathbf{CLR}_{dn} = \mathbf{Y}_{dn} / G_n \) uses the geometric mean \(G_n\) of \(\mathbf{Y}^{\parallel}_{n}\) to scale the data. They discovered that \(G_n\) was related to the scale of the underlying environment \( \mathbf{W\perp} \). Interestingly, the information in a system is a fundamental property and \(G_n\) is strongly correlated with Shannon's entropy, indicating that information about the complexity of \( \mathbf{W} \) is available post-sequencing (Supplemental). The realization that all normalizations made assumptions of scale led to the understanding that all normalizations  introduced latent bias into analyses because the scaling factor has been assumed up to now to be a perfect point estimate of the scale of the underlying system [@nixon2023scale].

With this new insight Nixon et al. [-@nixon2023scale] devised methods to set up a scale model that explicitly acknowledges and accounts for the uncertainty of the scale. In the case of the CLR which uses the  \(G_n\) as the denominator, this entails adding uncertainty into  \(G_n\) such that the latent bias can be exposed, acknowledged and accounted for.

The ALDEx2 R package [@fernandes:2013] represents a general purpose toolbox for Bayesian estimation of HTS datasets. ALDEx2 was designed originally to convert the discrete point estimates in the count matrix obtained into continuous posterior distributions through Monte-Carlo sampling from the Dirichlet distribution. The posterior more accurately represents variation on a per-sample basis[@gloorAJS:2016] and this \(\mathbf{Y}^{\parallel}_{dn}\) can more closely approximate \(\mathbf{W}^{\parallel}_{dn}\).  Each Monte-Carlo replicate is normalized using the CLR and used to calculate summary and test statistics, and finally  expected values and confidence intervals are reported across the Monte-Carlo replicates. The CLR calculation can be modified trivially to include a full scale model of the data by incorporating uncertainty into \(G\) and so produce a posterior estimate of the scale. The scale uncertainty  can be included on a per-experiment, per-group or even a per-sample basis. Thus, ALDEx2 can now be used to generate a full posterior model of both \(\mathbf{Y}^{\parallel}_{dn}\) and \(\mathbf{Y}^{\perp}_{dn}\).

@@

We start by examining the dispersion or variance of the data as counts and as the logarithm of the normalized counts as calculated by DESeq2 (log(RLE)) and by ALDEx2 (CLR). The actual counts of data derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010] as seen in Panel A of Figure 1. This relationship is why most tools model the counts with the Negative Binomial distribution[@polyester:2016], and use this distribution as the basis for batch correction[@Zhang:2020ab]. However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [@Robinson:2010; @Love:2014aa; @Andrews:2019aa], or on  the clr values [@fernandes:2013] both of which are log-ratios. The mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013; @Love:2014aa].

```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,  as logarithms of counts, and as clr values. Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale. Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first. Panel C shows the equivalent values calculated by ALDEx2  in which the expected clr value for each transcipt are plotted vs. the expected dispersion. The red line in each panel shows the loess line of fit to the mid-point of the distributions. In panels B and C the amount of dispersion reaches a minimum at moderate  values.  The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value. The values below the dashed grey line in panels B and C represent those below the first decile of dispersion." }

library(DESeq2)
devtools::load_all('~/Documents/0_git/ALDEx_bioc')
#library(ALDEx2)

@@

The very low dispersion estimate for most of the features  arises because  scale variation in the underlying data has been removed through sequencing and normalization. While the actual scale of the data is inaccessible post-sequencing  we can estimate the *effect* of scale on the output. To do this, we add uncertainty to the denominator that is used to calculate the log-ratio of the samples [@nixon2023scale], and combine this with the posterior probabilities that ALDEx2 calculates on a per-feature basis [@fernandes:2013]. Scale uncertainty is incorporated using the `gamma` parameter that controls the amount uncertainty being included when we call either `aldex()`, or `aldex.clr()`. The `ALDEx2` package further contains a sensitivity analysis function, `aldex.senAnalysis()`, that can be used to explore the effect of different amounts of scale uncertainty. In practice we suggest that a `gamma` parameter between 0.5 and 1 is realistic for most experimental designs.

@@

With this background, we can understand how to apply a full scale to the metatranscriptome dataset with the `aldex.makeScaleMatrix()` function. This function uses a logNormal distribution when building the model so that ratios that are chosen are symmetrical. The full scale model option can be used ot make a matrix of relative scales that are distinct for each group (or even each sample) along with their uncertainties. Applying a per-group relative differential scale of 0.15 with a base scale of 1 moves the housekeeping functions to the midline of no difference (Figure 3C).  Note that now a significant number of functions  are differentially up in BV that were formerly classed as not different without scale, or when only a uniform  scale was applied. These former false negatives are noted in orange in each panel. Inspection of the functions shows that these are largely missing from the Lactobacillus species and so should actually be captured as differentially abundant. Thus, applying differential scale allows us to distinguish between both false positives (housekeeping functions in cyan) and false negatives (orange functions) even in a very difficult to analyze dataset. We suggest that the default scale model alone is useful and should be normally applied when the data are well centred. However, when datasets are not well centred or when the investigator has prior information about the underlying biology, we advocate developing and using a full differential scale model.

@@

Since the underlying scale of the system cannot be measured easily -- methods such as flow cytometry [@Vandeputte:2017aa], spike-in probes and fluorescent in-situ hybridization [@Loven:2012aa;@yeast-absolute] have been used -- the effect on analysis can be modelled by including scale uncertainty in the analysis [@nixon2023scale]. Nixon et al, showed that this can be done by including uncertainty in the denominator used for the normalization. The ALDEx2 R package is ideally suited for this since this tool builds a Bayesian posterior of the compositional component of the dataset at the outset and then conducts the analysis on that posterior. Adding scale uncertainty can be done at the same time thus producing a posterior model that incorporates both compositional and scale uncertainty, and this can be done without additional computational complexity. For this, the compositional uncertainty is sampled from a Dirichlet distribution, and the scale uncertainty is sampled from a Normal or logNormal distribution depending on preference.

All normalizations attempt to make the samples in a dataset commensurate but do not explicitly address the scale of the underlying system. However, the general lack of scale information has important consequences for the analysis of HTS datasets. One issue is that analysis tools  seem over-powered with even moderate sample sizes and different tools have different intrinsic power, Type 1 and Type 2 error rates  [@Schurch:2016aa]. Using small sample sizes in analysis leads to less reliability and reproducibility in analyses since surprisingly large sample sizes are needed to determine reproducible p-values (eg. [@Halsey:2015aa]). Thus, recommendations to use small sample sizes in multivariate datasets such as RNA-seq datasets are not supported by simple modelling in the univariate case. Another issue is that datasets are difficult to analyze when there they contain systematic asymmetry, with different tools exhibiting differing pathologies with these datasets [@Robinson:2010a; @Wu2021].

>>>>

---
title: "Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2
"
shorttitle: "Scale ALDEx2"
author:
  - name: Greg Gloor
    affiliation: Department of Biochemistry, University of Western Ontario
    email: ggloor@uwo.ca
  - name: Michelle Pistner Nixon
    affiliation: College of Information Sciences and Technology, Pennsylvania	 State University
  - name: Justin Silverman
    affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania	 State University
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
output:
  nucleic-acids-research.csl: pdf_document
package: ALDEx2
abstract: 
  Despite scale being an important confounding variable, high throughput sequencing generates data that are difficult to reconcile with the scale of the underlying environment, but it has been shown that the scale of the environment is related to the geometric mean of the parts of the sample. The ALDEx2 R package builds a Bayesian model of the data and here we report an update to the model that includes both random sampling and scale estimates. We show that incorporating scale allows for robust inference in a number of different experimental designs and we provide guidance on how to incorporate scale inference into differential abundance analysis.  While all data normalizations in widespread use make implicit assumptions about the scale of the system we suggest that explicit rather than implicit scale models should be used.  
---

@@

High throughput sequencing (HTS) is a universal tool to explore many biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing). HTS proceeds by taking a fixed-number sample from the environment, making a library, multiplexing (merging) multiple libraries and applying a fixed number of molecules to the flow cell. In essence it is a poll of the environment that is mixed with other polls and then a poll of the mixture is taken. It should be clear that the total number of molecules sequenced is driven by the capacity of the instrument and not by the number of molecules in the sampled environment[@Lovell:2015;@Vandeputte:2017aa;@Props:2017aa] .

@@

More formally, if the true information in the  environment being sampled can be summarized by counts of features (genes, taxa, etc) in a matrix \( \mathbf{W} \), then the elements of the matrix can be decomposed into its composition (relative information, \( \mathbf{W}^{\parallel} \) ) and its scale (total sum, \( \mathbf{W}^{\perp} \) ); formally, \(\mathbf{W} = \mathbf{W}^{\parallel}  \mathbf{W}^{\perp}\). If the matrix has \(N\) samples and \(D\) parts, then we can uniquely identify \(\mathbf{W}_{dn}\) as the \(d^{th}\) feature in sample \(n\).  Any data that we collect by sequencing is thought to contain only proportional [@lovell:2011] information and we can represent the corresponding sequenced matrix as \(\mathbf{Y}\). Compositional data analysis makes the strong assumption	 that \(\mathbf{Y}_{n} \equiv \mathbf{Y}^{\parallel}_{n} = f(\mathbf{W}^{\parallel}_{n})\); in other words that what is measured by sequencing is relative information only, and that this measurement is a function of the relative information in the underlying environment. Note that under this  assumption no corresponding estimate or information is available for the scale of the sample \(\mathbf{Y}^{\perp}_{n}\).

@@

Recently, Nixon et al. [-@nixon2023scale] showed that that *all* normalizations make some  identifying assumption about the scale of \(\mathbf{W}\) but that the actual assumption being made was not always easy to interpret. In essence, normalizations in widespread use assume that either one sample can be chosen as a reference to which the others are scaled (TMM), or that different sub-parts of each sample maintain a constant scale across samples (RLE, CSS, CLR and derivatives).

The CLR, which can be written as \( \mathbf{CLR}_{(1...d)n} = \mathbf{Y}_{(1...d)n} / G_n \),  uses the geometric mean \( G_n \) of \( \mathbf{Y}^{\parallel}_{n} \) to scale the data. Nixon et al. [-@nixon2023scale] discovered that using the CLR was equivalent to making a statement that \( G_n \) was related to the scale of the underlying environment \( \mathbf{W}^{\perp} \). Interestingly, the information in a system is a fundamental property and \(G_n\) is strongly correlated with Shannon's entropy, indicating that information about the complexity of \( \mathbf{W} \) is available post-sequencing (Supplemental). The realization that all normalizations made assumptions of scale led to the understanding that all normalizations	 introduced latent bias into analyses because the scaling factor has been assumed up to now to be a perfect point estimate of the scale of the underlying system [@nixon2023scale].

With this new insight Nixon et al. [-@nixon2023scale] devised methods to set up a scale model that explicitly acknowledges and accounts for the uncertainty of the scale. In the case of the CLR which uses the	 \(G_n\) as the denominator, this entails adding uncertainty into  \(G_n\) such that the latent bias can be exposed, acknowledged and accounted for. Additionally it was realized that the actual values used in the denominator to calculate the log-ratio were irrelevant, but instead that the ratios between the denominators was the key parameter. These two insights allow for a full scale model of scale to be built and tested for its effect on analysis.

The ALDEx2 R package [@fernandes:2013] represents a general purpose toolbox for Bayesian estimation of HTS datasets. ALDEx2 was designed originally to convert the discrete point estimates in the count matrix obtained into continuous posterior distributions through Monte-Carlo sampling from the Dirichlet distribution. The posterior more accurately represents variation on a per-sample basis[@gloorAJS:2016]	 and this \(\mathbf{Y}^{\parallel}_{dn}\) can more closely approximate \(\mathbf{W}^{\parallel}_{dn}\).	 Each Monte-Carlo replicate is normalized using the CLR and used to calculate summary and test statistics, and finally	expected values and confidence intervals are reported across the Monte-Carlo replicates. The CLR calculation can be modified trivially to include a full scale model of the data by incorporating uncertainty into \(G\) and so produce a posterior estimate of the scale. The scale uncertainty  can be included on a per-experiment, per-group or even a per-sample basis. Thus, ALDEx2 can now be used to generate a full posterior model of both \(\mathbf{W}^{\parallel}_{dn}\) and \(\mathbf{W}^{\perp}_{dn}\) from the observed data.

@@

We start by examining the dispersion or variance of the data as counts and as the logarithm of the normalized counts as calculated by DESeq2 (log(RLE)) and by ALDEx2 (CLR). The actual counts of data derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010] as seen in Panel A of Figure 1. This relationship is why most tools model the counts with the Negative Binomial distribution[@polyester:2016], and use this distribution as the basis for batch correction[@Zhang:2020ab]. However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [@Robinson:2010; @Love:2014aa; @Andrews:2019aa], or on	the clr values [@fernandes:2013] both of which are log-ratios. The mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013; @Love:2014aa]. 
```{r, echo=F, warning=F, message=F,comment=F, result=F} 
library(DESeq2)
devtools::load_all('~/Documents/0_git/ALDEx_bioc')
#library(ALDEx2)
```
```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, cache=T, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,	 as logarithms of counts, and as clr values. Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale. Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first. Panel C shows the equivalent values calculated by ALDEx2	 in which the expected clr value for each transcipt are plotted vs. the expected dispersion. The red line in each panel shows the loess line of fit to the mid-point of the distributions. In panels B and C the amount of dispersion reaches a minimum at moderate	 values.  The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value. The values below the dashed grey line in panels B and C represent those below the first decile of dispersion." }

@@

The very low dispersion estimate for most of the features  arises because  scale variation in the underlying dataset has been removed through sequencing and normalization. While the actual scale of the data is inaccessible post-sequencing	we can estimate the *effect* of scale on the output. To do this, we add uncertainty to the denominator that is used to calculate the log-ratio of the samples [@nixon2023scale], and combine this with the posterior probabilities that ALDEx2 calculates on a per-feature basis [@fernandes:2013]. Scale uncertainty is incorporated using the `gamma` parameter that controls the amount uncertainty being included when we call either `aldex()`, or `aldex.clr()`. The `ALDEx2` package further contains a sensitivity analysis function, `aldex.senAnalysis()`, that can be used to explore the effect of different amounts of scale uncertainty. In practice we suggest that a `gamma` parameter between 0.5 and 1 is realistic for most experimental designs.

@@

With this background, we can understand how to apply a full scale to the metatranscriptome dataset with the `aldex.makeScaleMatrix()` function. This function uses a logNormal distribution when building the model so that ratios that are chosen are symmetrical. The full scale model option can be used to make a matrix of relative scales that are distinct for each group (or even each sample) along with their uncertainties. Applying a per-group relative differential scale of 0.15 with a base scale of 1 moves the housekeeping functions to the midline of no difference (Figure 3C).  Note that now a significant number of functions  are differentially up in BV that were formerly classed as not different without scale, or when only a uniform  scale was applied. These former false negatives are noted in orange in each panel. Inspection of the functions shows that these are largely missing from the Lactobacillus species and so should actually be captured as differentially abundant. Thus, applying differential scale allows us to distinguish between both false positives (housekeeping functions in cyan) and false negatives (orange functions) even in a very difficult to analyze dataset. We suggest that the default scale model alone is useful and should be normally applied when the data are well centred. However, when datasets are not well centred or when the investigator has prior information about the underlying biology, we advocate developing and using a full differential scale model.

@@

Since the underlying scale of the system cannot be measured easily -- methods such as flow cytometry [@Vandeputte:2017aa], spike-in probes and fluorescent in-situ hybridization [@Loven:2012aa;@yeast-absolute] have been used -- the effect on analysis can be modelled by including scale uncertainty in the analysis [@nixon2023scale]. Nixon et al, showed that this can be done by including uncertainty in the denominator used for the normalization. The ALDEx2 R package is ideally suited for this since this tool builds a Bayesian posterior of the compositional component of the dataset at the outset and then conducts the analysis on that posterior. Adding scale uncertainty can be done at the same time thus producing a posterior model that incorporates both compositional and scale uncertainty, and this can be done without additional computational complexity. For this, the compositional uncertainty is sampled from a Dirichlet distribution, and the scale uncertainty can be sampled from any distribution depending on prior knowledge or preference, but ALDEx2 samples from a logNormal distribution by default.

All normalizations attempt to make the samples in a dataset commensurate but do not explicitly address the scale of the underlying system. The two supplementary tables show that these normalizations can be interpreted as scale assumptions. However, the general lack of scale information has important consequences for the analysis of HTS datasets. One issue is that analysis tools	 seem over-powered with even moderate sample sizes and different tools have different intrinsic power, Type 1 and Type 2 error rates  [@Schurch:2016aa]. Using small sample sizes in analysis leads to less reliability and reproducibility in analyses since surprisingly large sample sizes are needed to determine reproducible p-values (eg. [@Halsey:2015aa]). Thus, recommendations to use small sample sizes in multivariate datasets such as RNA-seq datasets are not supported by simple modelling in the univariate case. Another issue is that datasets are difficult to analyze when there they contain systematic asymmetry, with different tools exhibiting differing pathologies with these datasets [@Robinson:2010a; @Wu2021].

Many groups have conducted benchmarking studies on different tools and normalizations used for the analysis of datasets such as transcriptomes [@Bullard:2010;@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa] and microbiomes[@Thorsen:2016aa;@Weiss:2017aa;@hawinkel2017]. Generically, it is observed  that the actual agreement between methods can be modest, and there is usually a recommendation that some methods are better suited for individual datasets or for specific types of analysis; e.g. 'Normalization and microbial differential abundance strategies depend upon data characteristics' [@Weiss:2017aa]. However, another interpretation of these studies is that the specific assumptions of the tools and the way they are used by different groups can lead to results that include both false negative and false positive results, skewed towards the latter. There are now published examples of multiple large replication efforts in psychology [@Open-Science-Collaboration:2015aa], cancer biology [@Rodgers:2021aa], ecology and evolution [@evoRepr-pre]. In a nutshell, experiences from these studies indicate that published data typically have higher effect sizes and appear more significant than replication efforts with the same data, or with newly generated replication data [@Rodgers:2021aa]. Thus, most published data should be viewed with suspicion [@Ioannidis:2005aa] unless efforts have been made to ensure that the results are robust
